---
title: "Matching in Practice: A Worked Example"
author: "Advanced Observational Causal Inference"
date: "September 9"
format:
  revealjs:
    smaller: true
    slide-number: true
    chalkboard: true
    incremental: false
    transition: none
    center: false
    width: 1280
    height: 720
    margin: 0.08
    preview-links: false
    html-math-method: mathjax
execute:
  echo: true
  warning: false
  message: false
---

## The teaching goal

We will use one synthetic dataset to see what matching actually does.

The sequence is:

$$
\text{naive comparison}
\rightarrow
\text{CEM}
\rightarrow
\text{nearest-neighbor matching}
\rightarrow
\text{weighting}
\rightarrow
\text{balance}.
$$

Because this is simulated data, we also know the true potential outcomes.

---

## Generate the data

```{r}
#| echo: true
#| output: false

library(tidyverse)
library(MatchIt)
library(cobalt)

set.seed(1234)

N <- 100

dat <- tibble(
  id = 1:N,
  age = pmin(pmax(rnorm(N, 40, 10), 20), 65),
  income = pmax(rnorm(N, 50 + 0.8 * age, 20), 5)
) |>
  mutate(
    college = rbinom(
      N, 1,
      plogis(-2 + 0.04 * age)
    ),

    propensity =
      plogis(
        -5.5 +
        0.055 * age +
        0.012 * income +
        0.9 * college
      ),

    D = rbinom(N, 1, propensity),

    Y0 =
      20 +
      0.35 * age +
      0.08 * income +
      5 * college +
      0.002 * income^2 +
      rnorm(N, 0, 5),

    tau = 8,

    Y1 = Y0 + tau,

    Y = D * Y1 + (1 - D) * Y0
  )
```

---

## What is the true ATT?

Because we simulated both potential outcomes, we can calculate the true ATT.

```{r}
true_ATT <- dat |>
  filter(D == 1) |>
  summarise(ATT = mean(Y1 - Y0)) |>
  pull(ATT)

true_ATT
```

---

## What does the naive comparison give?

```{r}
naive_ATT <- dat |>
  summarise(
    estimate =
      mean(Y[D == 1]) -
      mean(Y[D == 0])
  ) |>
  pull(estimate)

naive_ATT
```

The difference is larger than the true effect because the treated and control groups differ systematically on pre-treatment covariates.

---

## See the confounding

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5.5
#| out-width: 90%

ggplot(
  dat,
  aes(
    x = income,
    y = age,
    shape = factor(D)
  )
) +
  geom_point(size = 3, alpha = 0.7) +
  scale_shape_manual(
    values = c(19, 21),
    labels = c("Control", "Treated")
  ) +
  labs(
    x = "Income",
    y = "Age",
    shape = "",
    title = "Before matching"
  ) +
  theme_minimal(base_size = 14)
```

---

## Why not exact-match?

To exactly match on the original covariates, we would need

$$
age_i=age_j,\qquad
income_i=income_j,\qquad
college_i=college_j.
$$

With continuous `age` and `income`, this is essentially impossible.

So instead we **coarsen** the continuous variables.

---

## Create CEM strata

```{r}
dat <- dat |>
  mutate(
    age_band = cut(
      age,
      breaks = c(19, 29, 39, 49, 59, 69),
      labels = FALSE
    ),
    income_band = cut(
      income,
      breaks = c(0, 50, 75, 100, 125, Inf),
      labels = FALSE
    )
  )
```

A matching stratum is now defined by

$$
(age\ band,\ income\ band,\ college).
$$

---

## Visualize the strata

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5.5
#| out-width: 90%

ggplot(
  dat,
  aes(
    x = income,
    y = age,
    shape = factor(D)
  )
) +
  geom_vline(
    xintercept = c(50, 75, 100, 125),
    linetype = "dashed",
    alpha = 0.5
  ) +
  geom_hline(
    yintercept = c(29, 39, 49, 59),
    linetype = "dashed",
    alpha = 0.5
  ) +
  geom_point(size = 3, alpha = 0.7) +
  scale_shape_manual(
    values = c(19, 21),
    labels = c("Control", "Treated")
  ) +
  facet_wrap(
    ~ college,
    labeller = as_labeller(
      c(`0` = "No college", `1` = "College")
    )
  ) +
  labs(
    x = "Income",
    y = "Age",
    shape = "",
    title = "Coarsened exact matching",
    subtitle = "Each rectangle within a panel is a matching stratum"
  ) +
  theme_minimal(base_size = 14)
```

---

## Which strata survive?

```{r}
strata_status <- dat |>
  group_by(age_band, income_band, college) |>
  summarise(
    n_treated = sum(D == 1),
    n_control = sum(D == 0),
    .groups = "drop"
  ) |>
  mutate(
    retained = n_treated > 0 & n_control > 0
  )

strata_status |>
  count(retained)
```

A stratum survives only if it contains at least one treated and at least one control.

Strata without overlap are dropped.

---

## Show the dropped observations

```{r}
dat <- dat |>
  left_join(
    strata_status,
    by = c("age_band", "income_band", "college")
  ) |>
  mutate(
    matched = retained
  )
```

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5.5
#| out-width: 90%

ggplot(
  dat,
  aes(
    x = income,
    y = age,
    shape = factor(D)
  )
) +
  geom_point(
    data = filter(dat, matched),
    size = 3,
    alpha = 0.8
  ) +
  geom_point(
    data = filter(dat, !matched),
    size = 3,
    alpha = 0.2
  ) +
  scale_shape_manual(
    values = c(19, 21),
    labels = c("Control", "Treated")
  ) +
  labs(
    x = "Income",
    y = "Age",
    shape = "",
    title = "CEM drops strata without overlap",
    subtitle = "Faded observations lie in unusable strata"
  ) +
  theme_minimal(base_size = 14)
```

---

## The target population changes

Before matching:

$$
ATT_{full}
=
E[Y(1)-Y(0)\mid D=1].
$$

After dropping treated units with no matched controls:

$$
ATT_{matched}
=
E[Y(1)-Y(0)\mid D=1,\ matched=1].
$$
---

## The target population changes 

```{r}
ATT_full <- dat |>
  filter(D == 1) |>
  summarise(ATT = mean(Y1 - Y0)) |>
  pull(ATT)

ATT_matched <- dat |>
  filter(D == 1, matched) |>
  summarise(ATT = mean(Y1 - Y0)) |>
  pull(ATT)

tibble(
  quantity = c(
    "ATT: all treated",
    "ATT: matched treated"
  ),
  estimate = c(
    ATT_full,
    ATT_matched
  )
)
```

With a constant treatment effect, these should be nearly identical.

---

## Make treatment effects heterogeneous

Now let treatment effects vary with income:

$$
\tau_i
=
8+0.05(income_i-\bar{income}).
$$

```{r}
dat <- dat |>
  mutate(
    tau_het = 8 + 0.05 * (income - mean(income)),
    Y1_het = Y0 + tau_het,
    Y_het = D * Y1_het + (1 - D) * Y0
  )
```

---

## Trimming can now change the ATT

```{r}
ATT_full_het <- dat |>
  filter(D == 1) |>
  summarise(ATT = mean(Y1_het - Y0)) |>
  pull(ATT)

ATT_matched_het <- dat |>
  filter(D == 1, matched) |>
  summarise(ATT = mean(Y1_het - Y0)) |>
  pull(ATT)

tibble(
  quantity = c(
    "True ATT: all treated",
    "True ATT: matched treated"
  ),
  estimate = c(
    ATT_full_het,
    ATT_matched_het
  )
)
```

Now a difference can arise even though we know the potential outcomes exactly.

That difference is due to the change in the target population.

---

## Estimate the CEM ATT

```{r}
cem_ATT <- dat |>
  filter(matched) |>
  group_by(age_band, income_band, college) |>
  summarise(
    n_treated = sum(D == 1),
    n_control = sum(D == 0),
    treated_mean = mean(Y_het[D == 1]),
    control_mean = mean(Y_het[D == 0]),
    difference = treated_mean - control_mean,
    .groups = "drop"
  ) |>
  filter(n_treated > 0, n_control > 0) |>
  mutate(
    treated_weight = n_treated / sum(n_treated)
  ) |>
  summarise(
    ATT = sum(difference * treated_weight)
  ) |>
  pull(ATT)

cem_ATT
```

---

## Three different quantities

```{r}
tibble(
  quantity = c(
    "True ATT: all treated",
    "True ATT: matched treated",
    "Estimated CEM ATT"
  ),
  estimate = c(
    ATT_full_het,
    ATT_matched_het,
    cem_ATT
  )
)
```

The key distinction is:

$$
\widehat{ATT}_{CEM}-ATT_{full}
=
\underbrace{
ATT_{matched}-ATT_{full}
}_{\text{changed target}}
+
\underbrace{
\widehat{ATT}_{CEM}-ATT_{matched}
}_{\text{estimation error + residual imbalance}}.
$$

---

## Nearest-neighbor matching

Now avoid bins entirely.

Instead, define similarity continuously using Mahalanobis distance.

```{r}
m_nn <- matchit(
  D ~ age + income + college,
  data = dat,
  method = "nearest",
  distance = "mahalanobis",
  replace = TRUE,
  ratio = 1,
  estimand = "ATT"
)

summary(m_nn)
```

---

## Visualize nearest-neighbor matches

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5.5
#| out-width: 90%

pairs <- tibble(
  treated_id = as.numeric(rownames(m_nn$match.matrix)),
  control_id = as.numeric(m_nn$match.matrix[, 1])
) |>
  left_join(
    dat |> select(id, income, age,college),
    by = c("treated_id" = "id")
  ) |>
  rename(
    treated_income = income,
    treated_age = age
  ) |>
  left_join(
    dat |> select(id, income, age),
    by = c("control_id" = "id")
  ) |>
  rename(
    control_income = income,
    control_age = age
  )

ggplot(dat, aes(x = income, y = age, shape = factor(D))) +
  geom_point(size = 2.5, alpha = 0.35) +
  geom_segment(
    data = pairs,
    aes(
      x = control_income,
      y = control_age,
      xend = treated_income,
      yend = treated_age
    ),
    inherit.aes = FALSE,
    alpha = 0.5
  ) +
  scale_shape_manual(
    values = c(19, 21),
    labels = c("Control", "Treated")
  ) +
  labs(
    x = "Income",
    y = "Age",
    shape = "",
    title = "Nearest-neighbor matching"
  ) +
  theme_minimal(base_size = 14) + 
  facet_wrap(~college)
```

---

## Estimate the nearest-neighbor ATT

```{r}
nn_dat <- match.data(m_nn)

nn_fit <- lm(
  Y_het ~ D,
  data = nn_dat,
  weights = weights
)

nn_ATT <- coef(nn_fit)[["D"]]

nn_ATT
```

---

## ATT weighting

For the ATT, inverse-propensity weighting uses

$$
w_i =
\begin{cases}
~~~~~~~~ 1 & D_i=1,\\[4pt]
\dfrac{e(X_i)}{1-e(X_i)} & D_i=0.
\end{cases}
$$

This reweights the controls to resemble the treated population.

## IPW: the basic idea

Inverse probability weighting (IPW) does the same conceptual job as matching:

$$
\text{construct a comparison population}
$$

But instead of pairing or dropping observations, IPW changes how much each observation counts.

Let

$$
e(X_i)=P(D_i=1\mid X_i)
$$

be the propensity score.

---

## A simple example

Suppose Alice (who was treated) has

$$
e(X_A)=0.20
$$

Treatment was unusual for someone like Alice, so her ATE weight is

$$
w_A=\frac{1}{0.20}=5
$$

Now suppose Bob (who was also treated) has

$$
e(X_B)=0.80
$$

Treatment was expected for someone like Bob, so his ATE weight is

$$
w_B=\frac{1}{0.80}=1.25
$$

Alice counts more because her observed treatment state was less likely.

---

## The basic intuition

$$
\boxed{
\text{IPW upweights observations observed in an unlikely treatment state.}
}
$$

For treated units:

$$
w_i=\frac{1}{e(X_i)}.
$$

For untreated units:

$$
w_i=\frac{1}{1-e(X_i)}.
$$

The less likely the observed treatment status, the more informative that observation becomes for reconstructing the target population.

---

## Why does this help with confounding?

Suppose income affects treatment:

$$
income \rightarrow D
$$

and also affects the outcome:

$$
income \rightarrow Y.
$$

High-income people are disproportionately represented among the treated.

The raw data contain too many high-income treated units and too few high-income control units.

IPW changes their relative contributions.

---

## What happens to the treated group?

Suppose

$$
e(X)=0.9.
$$

A treated observation gets

$$
w=\frac{1}{0.9}\approx1.11.
$$

Now suppose

$$
e(X)=0.1.
$$

A treated observation gets

$$
w=\frac{1}{0.1}=10.
$$

Treated observations who were unlikely to be treated receive much more weight.

---

## What happens to the control group?

For controls, the relevant probability is

$$
1-e(X).
$$

Suppose

$$
e(X)=0.9
$$

but the observation is untreated.

Then

$$
w=\frac{1}{1-0.9}=10.
$$

This is a particularly informative control: someone who looks highly likely to have been treated, but happened not to be.

---

## IPW and the ATE

For the ATE,

$$
w_i=
\begin{cases}
\dfrac{1}{e(X_i)}, & D_i=1,\\[8pt]
\dfrac{1}{1-e(X_i)}, & D_i=0.
\end{cases}
$$

The resulting weighted groups are intended to resemble the same target population:

$$
\boxed{
\text{weighted treated}
\approx
\text{weighted controls}
\approx
\text{full population}.
}
$$

---

## The ATE estimator

A normalized IPW estimator is

$$
\widehat{ATE}_{IPW}
=
\frac{
\sum_i D_iY_i/e(X_i)
}{
\sum_i D_i/e(X_i)
}
-
\frac{
\sum_i(1-D_i)Y_i/[1-e(X_i)]
}{
\sum_i(1-D_i)/[1-e(X_i)]
}.
$$

The first term estimates the population mean under treatment.

The second estimates the population mean under control.

---

## IPW versus matching

The connection to matching is easiest to see for the ATT.

Matching gives us

$$
ATT
=
\sum_s
\Delta(s)P(S=s\mid D=1),
$$

where

$$
\Delta(s)
=
E[Y\mid D=1,S=s]
-
E[Y\mid D=0,S=s].
$$

We want the control group to have the **treated distribution of covariates**.

IPW accomplishes the same goal continuously, without explicitly constructing strata.

---

## ATT-IPW

For the ATT:

$$
w_i=
\begin{cases}
~~~~~~~~1, & D_i=1,\\[6pt]
\dfrac{e(X_i)}{1-e(X_i)}, & D_i=0.
\end{cases}
$$

The treated population is already our target population, so treated observations keep weight 1.

We only need to reweight the controls.

---

## ATT-IPW intuition

Consider three control observations:

$$
\begin{array}{c|c}
e(X) & \text{ATT weight}\\
\hline
0.10 & 0.11\\
0.50 & 1\\
0.90 & 9
\end{array}
$$

A control with $e(X)=0.90$ looks much more like the treated population.

So it gets much more weight when we construct

$$
E[Y(0)\mid D=1].
$$

---

## Why does the ATT weight have this form?

The control group has density proportional to

$$
f(X\mid D=0).
$$

The treated group has density proportional to

$$
f(X\mid D=1).
$$

By Bayes' rule,

$$
\frac{f(X\mid D=1)}{f(X\mid D=0)}
=
\frac{e(X)}{1-e(X)}
\cdot
\frac{P(D=0)}{P(D=1)}.
$$

The constant term does not matter after normalization.

Thus the ATT control weight is

$$
\boxed{
\frac{e(X)}{1-e(X)}.
}
$$

---


---

## What has the weighting accomplished?

Before weighting:

$$
f(X\mid D=1)
\neq
f(X\mid D=0).
$$

For the ATT, we want:

$$
f(X\mid D=1)
\approx
f_w(X\mid D=0).
$$

The weighting procedure creates a pseudo-control population that resembles the treated population.

---

## ATE versus ATT weighting

The target population determines the weights.

$$
\boxed{
\begin{array}{c|c|c}
& \text{Treated} & \text{Controls}\\
\hline
ATE &
1/e(X) &
1/[1-e(X)]\\[4pt]
ATT &
1 &
e(X)/[1-e(X)]
\end{array}
}
$$

ATE:

$$
\text{both groups} \rightarrow \text{full population}.
$$

ATT:

$$
\text{controls} \rightarrow \text{treated population}.
$$

---

## The overlap problem returns

Suppose

$$
e(X)=0.01
$$

Then its ATE weight is

$$
\frac{1}{0.01}=100.
$$

Likewise, if a control has

$$
e(X)=0.99,
$$

its ATE weight is also

$$
\frac{1}{1-0.99}=100.
$$

Lack of overlap appears as enormous weights.

---

## Matching and IPW handle lack of overlap differently

CEM says:

$$
\boxed{
\text{No comparable unit? Drop it.}
}
$$

Ordinary IPW says:

$$
\boxed{
\text{Comparable unit is very rare? Give it a large weight.}
}
$$

This is the same overlap problem appearing in two different forms.

---


## The problem with extreme weights

Extreme weights can make the estimate:

$$
\text{high variance}
$$

and can make the estimate depend heavily on a small number of observations.

This is why propensity-score overlap is not just a diagnostic issue.

It directly affects how informative the data are for the causal comparison.

---

## Overlap weights

A different strategy is to deliberately emphasize observations in regions of common support.

For overlap weights:

$$
w_i=
\begin{cases}
1-e(X_i), & D_i=1,\\[4pt]
e(X_i), & D_i=0.
\end{cases}
$$

Units with propensity scores near 0 or 1 receive little weight.

Units near

$$
e(X)=0.5
$$

receive the most weight.

---

## The broader picture

Matching and weighting are two ways of constructing a comparison population.

$$
\boxed{
\begin{array}{ccc}
\text{Matching}
&\rightarrow&
\text{construct comparable units}
\\[6pt]
\text{IPW}
&\rightarrow&
\text{construct comparable populations}
\\[6pt]
\text{Overlap weighting}
&\rightarrow&
\text{focus on common support}
\end{array}
}
$$

The identification assumption remains the same:

$$
(Y(1),Y(0))\perp D\mid X.
$$

The estimator determines how we use the observed data to exploit that assumption.


---

## Estimate ATT weights

```{r}
ps_model <- glm(
  D ~ age + income + college,
  data = dat,
  family = binomial()
)

dat <- dat |>
  mutate(
    ps = predict(ps_model, type = "response"),
    att_weight = if_else(
      D == 1,
      1,
      ps / (1 - ps)
    )
  )
```

---

## See the ATT weights

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5.5
#| out-width: 90%

ggplot(
  dat,
  aes(
    x = ps,
    y = att_weight,
    shape = factor(D)
  )
) +
  geom_point(size = 2.5, alpha = 0.7) +
  scale_shape_manual(
    values = c(19, 21),
    labels = c("Control", "Treated")
  ) +
  labs(
    x = "Estimated propensity score",
    y = "ATT weight",
    shape = "",
    title = "ATT inverse-propensity weights"
  ) +
  theme_minimal(base_size = 14)
```

---

## Estimate the weighted ATT

```{r}
att_ipw_fit <- lm(
  Y_het ~ D,
  data = dat,
  weights = att_weight
)

att_ipw <- coef(att_ipw_fit)[["D"]]

att_ipw
```

The weighted control group is now intended to represent the treated population.

---

## ATE weighting

For the ATE,

$$
w_i =
\begin{cases}
\dfrac{1}{e(X_i)} & D_i=1,\\[6pt]
\dfrac{1}{1-e(X_i)} & D_i=0.
\end{cases}
$$

This attempts to reconstruct the full target population.

---

## ATE weighting

```{r}
dat <- dat |>
  mutate(
    ate_weight = if_else(
      D == 1,
      1 / ps,
      1 / (1 - ps)
    )
  )

ate_ipw_fit <- lm(
  Y_het ~ D,
  data = dat,
  weights = ate_weight
)

ate_ipw <- coef(ate_ipw_fit)[["D"]]

ate_ipw
```

---

## Overlap weights

Overlap weights instead use

$$
w_i =
\begin{cases}
1-e(X_i) & D_i=1,\\
e(X_i) & D_i=0.
\end{cases}
$$

Units near the middle of the propensity-score distribution receive more weight.

```{r}
dat <- dat |>
  mutate(
    overlap_weight = if_else(
      D == 1,
      1 - ps,
      ps
    )
  )

overlap_fit <- lm(
  Y_het ~ D,
  data = dat,
  weights = overlap_weight
)

overlap_ate <- coef(overlap_fit)[["D"]]

overlap_ate
```

---

## Compare the weights visually

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5.5
#| out-width: 90%

ggplot(
  dat,
  aes(x = ps, y = overlap_weight, shape = factor(D))
) +
  geom_point(size = 2.5, alpha = 0.7) +
  scale_shape_manual(
    values = c(19, 21),
    labels = c("Control", "Treated")
  ) +
  labs(
    x = "Estimated propensity score",
    y = "Overlap weight",
    shape = "",
    title = "Overlap weights emphasize common support"
  ) +
  theme_minimal(base_size = 14)
```

---

## Balance is the diagnostic

Matching and weighting are procedures for constructing comparison populations.

We should therefore ask:

$$
\boxed{
\text{Did the procedure actually improve covariate balance?}
}
$$

```{r}
bal_nn <- bal.tab(
  m_nn,
  un = TRUE,
  m.threshold = 0.1
)

bal_nn
```

---

## Visualize balance

```{r}
#| output-location: slide
#| fig-width: 10
#| fig-height: 5.5
#| out-width: 90%

love.plot(
  bal_nn,
  stat = "mean.diffs",
  abs = TRUE,
  thresholds = c(m = 0.1),
  var.order = "unadjusted"
) +
  labs(
    title = "Covariate balance before and after matching"
  )
```

---

## The main lesson

Different estimators solve different parts of the problem.

$$
\begin{array}{c}
\text{Exact/CEM}
\rightarrow
\text{retain comparable strata}
\\[6pt]
\text{Nearest neighbor}
\rightarrow
\text{find similar units}
\\[6pt]
\text{Weighting}
\rightarrow
\text{reconstruct a target population}
\end{array}
$$

But in every case, we should distinguish:

$$
\boxed{
\text{What population is the estimator targeting?}
}
$$

from

$$
\boxed{
\text{How well did we construct that population?}
}
$$

and from

$$
\boxed{
\text{How much sampling/finite-sample error remains?}
}
$$
